---
title: "Overview: Data Formats and Analysis Requirements"
---

## The Core Question: Wide or Long?

Before diving into any analysis, you need to understand **data shape**. This is the single most important concept for preparing your data correctly.

### Wide Format

In **wide format**, each participant has ONE row, and repeated measurements appear in separate columns.

```{r}
#| echo: false
library(tidyverse)

# Example wide data
wide_example <- tibble(
  participant_id = c(1, 2, 3),
  age = c(22, 25, 21),
  condition = c("control", "treatment", "control"),
  score_time1 = c(45, 52, 48),
  score_time2 = c(47, 58, 51),
  score_time3 = c(44, 61, 53)
)

knitr::kable(wide_example, caption = "Wide format: One row per participant")
```

### Long Format

In **long format**, each observation has its own row. A participant with 3 measurements has 3 rows.

```{r}
#| echo: false
long_example <- wide_example |>
  pivot_longer(
    cols = starts_with("score_time"),
    names_to = "timepoint",
    values_to = "score",
    names_prefix = "score_"
  )

knitr::kable(long_example, caption = "Long format: One row per observation")
```

## Which Format for Which Analysis?

This is the critical insight that will save you hours of frustration:

| Analysis Type | Required Format | Why |
|--------------|-----------------|-----|
| Independent samples t-test | Wide | Comparing different groups (one row per person) |
| Paired/repeated t-test | Long (usually) | Comparing same person at different times |
| One-way between-subjects ANOVA | Wide | Comparing 3+ groups (one row per person) |
| One-way repeated measures ANOVA | Long | Same person, multiple conditions |
| Factorial between-subjects ANOVA | Wide | Multiple grouping variables |
| Mixed ANOVA | Long | Between and within factors combined |
| Multiple regression | Wide | Each predictor is a column |

::: {.callout-important}
## Key Insight
**Qualtrics exports data in WIDE format.** This is often (but not always!) what you need. The challenge is knowing when to convert to long format and how to do it correctly.
:::

## The Two Pathways

### Pathway 1: You Have Qualtrics Data

```{mermaid}
flowchart LR
    A[Qualtrics CSV] --> B[Remove metadata rows]
    B --> C[Clean column names]
    C --> D[Handle missing data]
    D --> E[Reverse score items]
    E --> F[Create composites]
    F --> G{Need long format?}
    G -->|Yes| H[Pivot to long]
    G -->|No| I[Ready for analysis]
    H --> I
```

**Challenges you'll face:**

1. Qualtrics adds 2-3 header rows with metadata
2. Column names are often long question text or codes like "Q4_1"
3. Response options may be text ("Strongly Agree") not numbers
4. Embedded data, timing, and metadata columns clutter the dataset

### Pathway 2: Designing Optimal Data from Scratch

If you can control how data is collected (or simulated), structure it optimally from the start:

**For between-subjects designs:**
```
participant_id, group, dv
1, control, 45
2, treatment, 52
3, control, 48
```

**For within-subjects designs:**
```
participant_id, condition, dv
1, baseline, 45
1, post, 52
2, baseline, 48
2, post, 55
```

**For factorial designs:**
```
participant_id, factor_a, factor_b, dv
1, low, morning, 45
2, low, evening, 52
3, high, morning, 48
4, high, evening, 55
```

## Data Requirements by Analysis

### t-test Requirements

**Independent samples t-test:**

- One grouping variable (2 levels)
- One continuous dependent variable
- Each participant appears once
- Format: Wide

```{r}
#| echo: false
ttest_indep <- tibble(
  participant_id = 1:6,
  group = rep(c("control", "treatment"), each = 3),
  score = c(45, 48, 42, 58, 62, 55)
)
knitr::kable(ttest_indep)
```

**Paired samples t-test:**

- Same participants measured twice
- Format: Can be wide OR long (depends on function used)

```{r}
#| echo: false
# Wide version
ttest_paired_wide <- tibble(
  participant_id = 1:4,
  pre_score = c(45, 48, 42, 50),
  post_score = c(52, 55, 48, 58)
)
knitr::kable(ttest_paired_wide, caption = "Paired t-test: Wide format")
```

### ANOVA Requirements

**One-way between-subjects ANOVA:**

- One grouping variable (3+ levels)
- One continuous DV
- Format: Wide

```{r}
#| echo: false
anova_oneway <- tibble(
  participant_id = 1:9,
  group = rep(c("control", "low_dose", "high_dose"), each = 3),
  score = c(45, 48, 42, 52, 55, 50, 62, 65, 58)
)
knitr::kable(anova_oneway)
```

**Repeated measures ANOVA:**

- Same participants across all conditions
- Format: Long

```{r}
#| echo: false
anova_rm <- tibble(
  participant_id = rep(1:3, each = 3),
  timepoint = rep(c("baseline", "week4", "week8"), 3),
  score = c(45, 52, 58, 48, 54, 61, 42, 50, 55)
)
knitr::kable(anova_rm)
```

**Factorial ANOVA (2x2 between-subjects):**

- Two grouping variables
- Format: Wide

```{r}
#| echo: false
anova_factorial <- tibble(
  participant_id = 1:8,
  drug = rep(c("placebo", "active"), each = 4),
  therapy = rep(c("none", "CBT", "none", "CBT"), 2),
  score = c(45, 52, 48, 58, 55, 68, 52, 72)
)
knitr::kable(anova_factorial)
```

### Multiple Regression Requirements

- Multiple predictor variables (continuous or categorical)
- One continuous outcome variable
- Format: Wide (one row per participant)

```{r}
#| echo: false
regression_example <- tibble(
  participant_id = 1:6,
  age = c(22, 25, 21, 28, 24, 26),
  hours_study = c(10, 15, 8, 20, 12, 18),
  motivation = c(4.2, 4.8, 3.5, 4.9, 4.0, 4.5),
  test_score = c(72, 85, 65, 92, 78, 88)
)
knitr::kable(regression_example)
```

## Common Qualtrics Scenarios

### Scenario 1: Simple Survey (Between-Subjects)

You've surveyed participants, randomly assigning them to view one of two advertisements, then measuring their purchase intention.

**Qualtrics gives you:** Wide format (perfect!)
**You need:** Clean it, maybe create composite scores, then analyse

### Scenario 2: Pre-Post Design

Participants complete a baseline measure, undergo an intervention, then complete the same measure again.

**Qualtrics gives you:** Wide format with `measure_pre` and `measure_post` columns
**You need:** Convert to long format for repeated measures analysis

### Scenario 3: Multiple Trials

Participants respond to 10 different stimuli, rating each on several dimensions.

**Qualtrics gives you:** Wide format with 10 × dimensions columns
**You need:** Convert to long format where each trial is a row

## Summary: The Decision Tree

```{mermaid}
flowchart TD
    A[What analysis are you running?] --> B{Same participants<br/>measured multiple times?}
    B -->|No| C[Keep Wide Format]
    B -->|Yes| D{Using base R t.test<br/>with paired=TRUE?}
    D -->|Yes| E[Can stay Wide]
    D -->|No| F[Convert to Long]
    C --> G[Clean and Analyse]
    E --> G
    F --> G
```

## Next Steps

In the following chapters, we'll work through:

1. **Cleaning Qualtrics exports** - Removing junk, fixing names, handling metadata
2. **Understanding optimal formats** - What your data should look like
3. **Reshaping data** - Converting between wide and long
4. **Reverse scoring** - Handling negatively-worded items
5. **Computing composites** - Creating scale scores and checking reliability
6. **Preparing for each analysis** - Specific requirements for t-tests, ANOVA, regression
