---
title: "Creating Composite Scores and Checking Reliability"
---

## Overview

After reverse scoring your items, the next step is to create composite scores (scale means or totals) and verify the internal consistency of your measures using Cronbach's alpha.

::: {.callout-note}
## Chapter Status: Placeholder
This chapter is under development. It will cover:

- Calculating scale means vs. totals
- Using `rowMeans()` and `rowSums()`
- Handling missing data in composites
- Cronbach's alpha with the `psych` package
- Interpreting alpha values
- Item-total correlations
- When to drop items
:::

## Topics to Be Covered

### Creating Composite Scores

```{r}
#| eval: false
# Mean composite (more common)
data <- data |>
  rowwise() |>
  mutate(
    wellbeing_mean = mean(c(wb_1, wb_2_r, wb_3, wb_4_r, wb_5), na.rm = TRUE)
  ) |>
  ungroup()

# Sum composite
data <- data |>
  rowwise() |>
  mutate(
    wellbeing_total = sum(c(wb_1, wb_2_r, wb_3, wb_4_r, wb_5), na.rm = TRUE)
  ) |>
  ungroup()
```

### Checking Reliability

```{r}
#| eval: false
library(psych)

# Select scale items (with reverse-scored versions)
scale_items <- data |>
  select(wb_1, wb_2_r, wb_3, wb_4_r, wb_5)

# Calculate alpha
alpha_result <- alpha(scale_items)
print(alpha_result)
```

### Interpreting Alpha

| Alpha Value | Interpretation |
|-------------|----------------|
| > 0.90 | Excellent |
| 0.80 - 0.89 | Good |
| 0.70 - 0.79 | Acceptable |
| 0.60 - 0.69 | Questionable |
| < 0.60 | Poor |

## Coming Soon

Full worked examples with:

- Multiple scales in one dataset
- Decision rules for item removal
- Reporting reliability in APA format
