---
title: "Preparing Data for t-tests"
---

## Overview

This chapter covers the final preparation steps specific to t-tests, including checking assumptions and structuring data for both independent and paired samples designs.

::: {.callout-note}
## Chapter Status: Placeholder
This chapter is under development. It will cover:

- Independent samples t-test data requirements
- Paired samples t-test data requirements
- Wide vs. long format considerations
- Assumption checking (normality, homogeneity of variance)
- Complete worked examples
:::

## Topics to Be Covered

### Independent Samples t-test

**Data Format Required:** Wide (one row per participant)

```{r}
#| eval: false
# Data structure
# participant_id | group | outcome
# 1              | control | 45
# 2              | treatment | 52

# Run the test
t.test(outcome ~ group, data = my_data)
```

### Paired Samples t-test

**Data Format Options:**

1. **Wide format** (works with base R)

```{r}
#| eval: false
# participant_id | pre | post
t.test(my_data$pre, my_data$post, paired = TRUE)
```

2. **Long format** (works with modern packages)

```{r}
#| eval: false
# participant_id | time | score
library(rstatix)
my_data |> t_test(score ~ time, paired = TRUE)
```

### Assumption Checking

```{r}
#| eval: false
# Normality
shapiro.test(my_data$outcome)

# Homogeneity of variance
car::leveneTest(outcome ~ group, data = my_data)
```

## Coming Soon

- Complete workflow from Qualtrics to t-test
- Effect size calculation (Cohen's d)
- Visualising group differences
- Handling assumption violations
- Multiple trials/items within conditions
