---
title: "Preparing Data for ANOVA"
---

## Overview

This chapter covers data preparation for different ANOVA designs, from simple one-way to complex mixed designs.

::: {.callout-note}
## Chapter Status: Placeholder
This chapter is under development. It will cover:

- One-way between-subjects ANOVA
- One-way repeated measures ANOVA
- Factorial (2×2, 2×3, etc.) ANOVA
- Mixed ANOVA (between × within)
- Assumption checking
- Post-hoc tests
:::

## Topics to Be Covered

### One-way Between-Subjects ANOVA

**Data Format:** Wide

```{r}
#| eval: false
# participant_id | group | outcome
model <- aov(outcome ~ group, data = my_data)
summary(model)
```

### One-way Repeated Measures ANOVA

**Data Format:** Long

```{r}
#| eval: false
# participant_id | condition | score
library(afex)
aov_ez(id = "participant_id",
       dv = "score",
       data = my_data,
       within = "condition")
```

### Factorial ANOVA

**Data Format:** Wide (between-subjects factors)

```{r}
#| eval: false
# participant_id | factor_a | factor_b | outcome
model <- aov(outcome ~ factor_a * factor_b, data = my_data)
summary(model)
```

### Mixed ANOVA

**Data Format:** Long (for within-subjects factor)

```{r}
#| eval: false
# participant_id | between_factor | within_factor | score
library(afex)
aov_ez(id = "participant_id",
       dv = "score",
       data = my_data,
       between = "between_factor",
       within = "within_factor")
```

## Coming Soon

- Assumption testing (sphericity, homogeneity)
- Effect sizes (eta-squared, partial eta-squared)
- Post-hoc comparisons with corrections
- Visualising ANOVA results
- Reporting in APA format
