---
title: "Preparing Data for Multiple Regression"
---

## Overview

This chapter covers data preparation for multiple regression analysis, including handling categorical predictors and covariates.

::: {.callout-note}
## Chapter Status: Placeholder
This chapter is under development. It will cover:

- Data format for regression (always wide)
- Continuous vs. categorical predictors
- Coding categorical variables
- Including covariates
- Centering and standardising predictors
- Assumption checking
:::

## Topics to Be Covered

### Basic Multiple Regression

**Data Format:** Wide (one row per participant, predictors as columns)

```{r}
#| eval: false
# participant_id | predictor1 | predictor2 | outcome
model <- lm(outcome ~ predictor1 + predictor2, data = my_data)
summary(model)
```

### Categorical Predictors

```{r}
#| eval: false
# Ensure proper factor coding
my_data <- my_data |>
  mutate(
    group = factor(group, levels = c("control", "treatment")),
    group = relevel(group, ref = "control")  # Set reference level
  )

model <- lm(outcome ~ group + age + education, data = my_data)
```

### Including Covariates (ANCOVA-style)

```{r}
#| eval: false
# Control for pre_score when examining treatment effect
model <- lm(post_score ~ treatment + pre_score, data = my_data)

# Type III sums of squares
library(car)
Anova(model, type = 3)
```

### Centering Predictors

```{r}
#| eval: false
my_data <- my_data |>
  mutate(
    age_centered = age - mean(age, na.rm = TRUE),
    age_z = scale(age)[,1]  # Standardised (z-score)
  )
```

## Coming Soon

- Assumption checking (linearity, normality, homoscedasticity)
- Multicollinearity (VIF)
- Effect sizes (R², adjusted R², f²)
- Interaction terms
- Model comparison
- Reporting in APA format
