---
title: "Shiny Data Converter Tool"
---

## Overview

A simple Shiny app to help convert between wide and long formats without writing code.

::: {.callout-note}
## Chapter Status: Placeholder
This chapter will include:

- Interactive Shiny app for data conversion
- Upload CSV, select conversion type
- Preview and download converted data
- Code snippets generated for reproducibility
:::

## Planned Features

### Upload and Preview

- Upload CSV file
- Automatic detection of potential ID columns
- Preview first 10 rows

### Wide to Long Conversion

- Select columns to pivot
- Name the new grouping variable
- Name the new values variable
- Preview result before download

### Long to Wide Conversion

- Select the names_from column
- Select the values_from column
- Preview result before download

### Code Generation

The app will generate R code so you can reproduce the transformation:
```{r}
#| eval: false
# Generated code:
library(tidyverse)

data_long <- data |>
  pivot_longer(
    cols = c(score_t1, score_t2, score_t3),
    names_to = "timepoint",
    values_to = "score",
    names_prefix = "score_"
  )
```

## Coming Soon

Interactive Shiny app embedded in this document or available as standalone application.

```{r}
#| eval: false
# To run the app (once developed):
# shiny::runApp("R/shiny_converter")
```

## Quick Reference: Code Snippets

While the Shiny app is being developed, here are common conversion patterns:

### Wide to Long

```{r}
#| eval: false
# Basic
data |>
  pivot_longer(
    cols = starts_with("score_"),
    names_to = "time",
    values_to = "score"
  )

# With prefix removal
data |>
  pivot_longer(
    cols = starts_with("score_"),
    names_to = "time",
    values_to = "score",
    names_prefix = "score_"
  )
```

### Long to Wide

```{r}
#| eval: false
# Basic
data |>
  pivot_wider(
    names_from = time,
    values_from = score
  )

# With prefix
data |>
  pivot_wider(
    names_from = time,
    values_from = score,
    names_prefix = "score_"
  )
```
